A Large-scale Study of Polymorphism in R

### 2019-03-29 notebooks/corpus_stats_and_plots.Rmd
merged airplane changes. rip wow air

total_recorded_sigs %>%
  inner_join(records_by_fun, by="names") %>%
  filter(obs != 0, rec !=0) %>% 
  inner_join(cloc, by="names") %>%
  filter(rec <= obs) %>%                      # Added this
  filter(r_counts != 0) -> df

@@

Similar to the R stuff.
Now, we should combine these plots.
Messy graphs, beware!
```{r}
loc_of_top_10 <- lapply(1:length(combo_df$package_name), function(x) if (combo_df$package_name[[x]] %in% top_10) x else FALSE)

@@

zzz %>%
ggplot(aes(x=names_o, value, group=1)) +
 geom_line(aes(colour = language), size=0.008) +
 labs(title = "Analysis of R Corpus (R and C)", y = "Lines of Code in Package (Log10-scaled)") + guides(colour=F) + 
 theme( axis.text.x  = element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank(),
        axis.ticks.y = element_line(),  panel.grid.major.y = element_line(colour = "red")) + # ,
  scale_y_continuous(name="Lines of Code in Package", breaks = c(-4, -3, -2, -1, 0, 1, 2, 3, 4), 
                     labels = c(10000, 1000, 100, 10, 0, 10, 100, 1000, 10000))

@@

ggsave("../paper/plots/linesofrandccode.pdf")
```

Plot:
```{r}
df_rc %>%
  mutate(rec=log(rec),obs=log(obs))  %>%
  ggplot(aes(x=rec, y=obs)) +
                geom_point(size=0.1) + 
                labs( title = "Numbers of Observed Signatures",
                      y = "Obs Invocations in Pkg",
                      x = "Rec Invocations in Pkg")
                # theme(axis.text.x = element_blank(), axis.title.x=element_blank(),
                #       axis.ticks.x=element_blank()) #+
                # scale_y_continuous(breaks = c(0, 1, 2, 3, 4, 5, 6),
                #      labels = c(0, 10, 100, 1000, 10000, 100000, 1000000))

# generate file
ggsave("../paper/plots/recordsbypkg.pdf")
```

>>>>

total_recorded_sigs %>%
  inner_join(records_by_fun, by="names") %>%
  filter(obs != 0, rec !=0) %>% 
  inner_join(cloc, by="names") %>%
  filter(rec <= obs) %>%                      # Added this: FIXME, understand why
  filter(r_counts != 0) -> df

@@

Similar to the R stuff.
Now, we should combine these plots.
Messy graphs, beware!
```{r}
#loc_of_top_10 <- lapply(1:length(combo_df$package_name), function(x) if (combo_df$package_name[[x]] %in% top_10) x else FALSE)

@@

zzz %>%
ggplot(aes(x=names_o, value, group=1)) +
 geom_line(aes(colour = language), size=0.008) + guides(colour=F) + 
 theme( axis.text.x  = element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank(),
        axis.ticks.y = element_line(),  panel.grid.major.y = element_line(colour = "red")) + # ,
  scale_y_continuous(name="", breaks = c(-4, -3, -2, -1, 0, 1, 2, 3, 4), 
                     labels = c(10000, 1000, 100, 10, 0, 10, 100, 1000, 10000))

@@

ggsave("../paper/plots/linesofrandccode.pdf",width=10,height=4)
```

Plot:
```{r}
df_rc %>%
  mutate(rec=log10(rec),obs=log10(obs))  %>%
  ggplot(aes(y=rec, x=obs)) + geom_point(size=0.1) + 
    labs( x = "Observed", y = "Recorded") +
    scale_x_continuous(breaks = c( 1, 2, 3, 4, 5, 6, 7),
                       labels = c( 10, 100, 1000, 10000, 100000, 1000000, 10000000))+
    scale_y_continuous(breaks = c( 1, 2, 3, 4, 5, 6, 7),
                       labels = c( 10, 100, 1000, 10000, 100000, 1000000, 10000000))

# generate file
ggsave("../paper/plots/recordsbypkg.pdf",width=9,height=5)
```

### 2019-03-29 paper/main.tex
merged airplane changes. rip wow air

In this section, we present our dataset. R aims to accommodate data
analysts, their workflows start with data import, followed by cleaning, and
then by steps of modeling, transformation and visualization. Often, the code
of these analysis pipelines resides, together with the data and results, in
notebooks. Few notebooks are publicly shared, and when they are, the data
isn't. For this reason our analysis focuses on packages which bundle
reusable units of R code with documentation, sample data and use-cases.

We focus on packages hosted on the \emph{Comprehensive R Archive Network} or
CRAN.  With over 13,000 packages, CRAN is the largest repository of software
written in R. It is experiencing sustained growth with an average of size
new packages a day~\cite{LIgges2017}.  Unlike sites like GitHub, CRAN is a
curated collection of packages: A package is only accepted to CRAN if it
abides by a number of well-formedness rules.  Most relevant for our
purposes, packages must have data, examples, vignettes and tests, all of
which must successfully run. From our perspective this means that each
package in CRAN comes with several executable scripts that exercise some of
its functionality.  Notable exceptions to this rule are packages only
containing data, which have no runnable code but are referenced by other
packages.
In our experience, only \DATAPKGS packages had no executable test or example code,
accounting for \DATAPKGSPERC of CRAN.

Our corpus is a subset of CRAN, consisting of \PACKAGES packages, accounting
for some \PERCENTCRAN of all packages.  These packages have a total of \RLOC
lines of R code and \CLOC lines of C code. For each package, we extracted
all executable code snippets from documentation, vignettes and tests and ran
them independently recording all calls to R functions.  It is noteworthy that
in order to run the scripts in one package, it is often necessary to load a
number of other packages.  In~\cite{issta18}, the authors estimated code
coverage to be around 68\% when including reverse dependencies.  As our infrastructure
adds overhead to script execution, coupled with the fact that some scripts take 
inordinate amounts of time to run, we limit our analysis of any given package to one hour.
We capture and record function signatures on-the-fly, though, so we do collect (partial) data even 
for these long-running packages.

Figure~\ref{most} shows the ten most downloaded CRAN packages.  For each
one, we list how many lines of R and C/C++ the packages contains.
Typically, C/C++ code is used to implement performance critical portions of
the code. We show the number of scripts that could be extracted from the
package. Each script corresponds to either one use-case or a set of unit
tests.

Figure~\ref{allcloc} shows the size of the code across all the packages we have used. 
The lines of R code are above the X-axis, and lines of C code below.
All counts are $log_{10}$-scaled, and entries are in ascending order by number of lines of R.
Roughly speaking, the majority of R packages have between 100 and 1000 lines of code.
In terms of C, only ~25\% of packages have any C code.
C and C++ are often used to implement parts of a package where performance is critical.

Lastly, Figure~\ref{recorded} plots the relationship between recorded and observed signatures.
Here, \# observed $>$ \# recorded since it's impossible to record more traces than those which pass through the tracer.
Numbers are again $log_{10}$-scaled.

\AT{TODO Touch up graphs.}
\AT{TODO Other interesting plots? Recorded vs \# scripts?}

Having built a sense for the corpus of data that we're dealing with, we will turn our attention to our methods of data collection.

\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=0.6]{linesofrandccode}
\caption{Lines of code in the analyzed corpus. \# of lines of R code are above 0, \# of lines of C code are below 0.}
\label{allcloc}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=0.6]{recordsbypkg}
\caption{Numbers of Recorded Function Invocations in the Analyzed Corpus}
\label{recorded}
\end{center}
\end{figure}

@@

@@

In this section, we will detail our methodology for collecting data.  The
basic idea we are exploring is to observe arguments and return values of
function calls, and from these generalize possible type signatures for the
called functions. To this end, we build on an open source tool called
\genthat whose purpose is to generate unit tests for R
libraries~\cite{issta18}.  To generate test cases, one must record and
accurately recreate arguments and return values and avoid generating
redundant tests (i.e. tests that do not increase code coverage). For our
purposes, we change the existing tool in two main ways: We record argument
\emph{shapes} rather than their values, and given the limited number of shapes
this allowed us to avoid having to run the code coverage minimization phase of
the tool. Both changes were beneficial for scalability---allowing us to
trace more calls than was reported in~\cite{issta18}.

@@

To illustrate our approach, consider the following script which adds a double
to an integer, and then creates a matrix and a vector, and adds them together to
get a new matrix.

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
> 1 + 2L
[1] 3
> x <- c(1,2,3,4)
> attr(x,"dim") <- c(2,2)
> x + c(1,2)
     [,1] [,2]
[1,]    2    4
[2,]    4    6
\end{lstlisting}}\caption{Example script}\label{example}\end{figure}

>>>>

In this section, we present our dataset. The R language aims to accommodate
data analysts; their workflows start with data import, followed by cleaning,
and then by steps of modeling, transformation and visualization. Often, the
code of these analysis pipelines resides, together with the data and
results, in notebooks. Few notebooks are publicly shared, and when they are,
the data isn't. For this reason our analysis focuses on packages which
bundle reusable units of R code with documentation, sample data and
use-cases.

We focus on packages hosted on the \emph{Comprehensive R Archive Network} or
CRAN.  With over 13,000 packages, CRAN is the largest repository of software
written in R. It is experiencing sustained growth with an average of size
new packages a day~\cite{LIgges2017}.  Unlike sites like GitHub, CRAN is a
\emph{curated} collection: A package is only accepted to CRAN if it abides
by a number of well-formedness rules.  Most relevant for our purposes,
packages must have data, examples, vignettes and tests, all of which must
successfully run. From our perspective this means that each package in CRAN
comes with several executable scripts that exercise some of its
functionality.  Notable exceptions to this rule are packages only containing
data, which have no runnable code but are referenced by other packages.
Only \DATAPKGS packages had no executable code, accounting for \DATAPKGSPERC
of CRAN.

The corpus used in this paper is a subset of CRAN. We retained packages that
could be run by our infrastructure in less than one hour. This corpus
consists of \PACKAGES packages, accounting for some \PERCENTCRAN of all
packages.  These packages have a total of \RLOC lines of R code and \CLOC
lines of C code. Figure~\ref{allcloc} shows a per-package breakdown of the
size of each package sorted by increasing numbers of lines of R. The figure
suggests that there is little correlation between the C/C++ parts and how
many lines of R a package contains. The median size of a package is 541
lines of R code and the largest package has 86K LOC. Typically, C/C++ code
is used to implement performance critical portions of the code. The majority
of package, 8,375 to be precise, have no C or C++ code.  The remaining 3,078
packages, have a median 572 lines of C and the largest package has 385,839
lines of C.

\begin{figure}[!b]\begin{center}
\includegraphics[width=.9\textwidth]{linesofrandccode}
\caption{Lines of code (log scal); for each package, R is above 0, and C/C++
  below}\label{allcloc}\end{center}
\end{figure}

For each package, we extracted all executable code snippets from
documentation, vignettes and tests and ran them independently recording all
calls to R functions.  It is noteworthy that in order to run the scripts in
one package, it is often necessary to load a number of other packages.
In~\cite{issta18}, the authors estimated code coverage to be around 68\%
when including reverse dependencies.  As our infrastructure adds overhead to
script execution, coupled with the fact that some scripts take inordinate
amounts of time to run, this is why we limited our analysis of any given
package to one hour.

@@

Figure~\ref{most} shows the ten most downloaded CRAN packages.  For each
one, we list how many lines of R and C/C++ the packages contains.  We show
the number of scripts that could be extracted from the package. Each script
corresponds to either one use-case or a set of unit tests. We print the
number of function calls that were observed by our infastructure, and the
number of unique signatures that were recorded.

Our infrastructure only retains unique argument/return combinations. Thus,
while we observe large number of functions being called with different
values, the types of these function call are often similar. Over the entire
corpus, we can see the relation between observed and recorded calls in
Figure~\ref{recorded}.  The median number of observed calls is 82 and
maximum is 19 million.  The median number of recorded signatures is 16 and
the maxium is 8,422. These numbers are skewed by a number of scripts doing
very few calls before plunging into C code.

\begin{figure}[htbp]\begin{center}
\includegraphics[width=.9\textwidth]{recordsbypkg}
\caption{Numbers of Recorded Function Invocations in the Analyzed Corpus}
\label{recorded}\end{center}
\end{figure}

@@

In this section, we will detail our methodology for collecting data.  
Basically, our aim is to observe arguments and return values of function calls, and from these generalize possible type signatures for the called functions. 
To this end, we build on an open source tool called \genthat whose purpose is to generate unit tests for R libraries~\cite{issta18}.  
\genthat achieves this by synthesizing unit tests from recorded function argument and return values, comparing the test against existing ones to avoid generating tests which do not increase code coverage.
To suit our purposes, we change the existing tool in two main ways: 
We record argument \emph{shapes} rather than values, and given the limited number of shapes this allowed us to avoid having to run the code coverage minimization phase of the tool. 
Both changes were beneficial for scalability, allowing us to trace far more calls than was reported in the \genthat work~\cite{issta18}.

@@

To illustrate our approach, consider the following script which adds a double to an integer, and then creates a matrix from a vector, and adds them together to get a new matrix.

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
> 1 + 2L                        # add double and integer
[1] 3
> x <- c(1,2,3,4)               # save double vector
> attr(x,"dim") <- c(2,2)       # assign to "dim" attribute
> x + c(1,2)                    # add double 3 to each matrix element matrix
     [,1] [,2]
[1,]    2    4
[2,]    4    6
\end{lstlisting}}\caption{Example script.}\label{example}\end{figure}
